---
title: "Machine Learning and Predictive Modeling"
author: 
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    toc_depth: '3'
  pdf_document:
    toc: true
    toc_depth: 3
    number_sections: true
    latex_engine: xelatex
fontsize: 8pt
subtitle: Forecasting US Federal Reserve Policy.
---
# Introduction

The United States Federal Reserve (**Fed**) is the organisation that is responsible for the monetary policy of the USA and is considered an **independent** government agency(central bank) within the federal system.

The Fed as a central bank has the capacity to **issue** coin and paper currency and also redeems various types of securities (bonds, bills, notes). It came into existence as an entity that would theoretically be able to address financial systemic risk and as such processes all necessary tools and regulatory oversight to do so. 

Aside from money printing and redeeming securities it also functions as a **primary lender**. The provided liquidity flows mostly to institutions with depository function, like **banks**. In special cases like **crises** the Fed is allowed to lend to the **private** and **business** sectors as well. [Section13(3)FRA]

The lending rate is a multivariate problem that the fed is constantly adjusting and monitoring based on indicators related to the health of the US economy.

It is important to note that the Fed's main policy committee meets every six weeks and makes a policy statement but does not always adjust rates. **Adjustment is dependent on necessity.**

Which in turn is dependent on specific variables or metrics the that the Fed is considerate about.

We will briefly describe the specific metrics and evaluate them as **explanatory variables** (factors) for our forecasting approach.

**Since the abolition of the gold standard the dollar itself has become the backbone of the current worldwide economic status quo. Thus the policy applied by the Federal Reserve has far reaching implications. This is the primary motivation for this report.**

# Methodology

## Variables

### Interest Rates (Treasuries)

- **DGS2**

2-Year Treasury Constant Maturity Rate Yield on a 2-year U.S. Treasury security. Very sensitive to Fed policy expectations.
Often used to gauge short-term rate outlook

- **DGS10**

10-Year Treasury Constant Maturity Rate. Benchmark long-term U.S. interest rate
Used in mortgages, valuation models, discounting. Core input for term spread (10Y – 2Y)

### Corporate Credit

- **BAMLC0A1CAAAEY**

Yield on AAA-rated corporate bonds. Measures cost of borrowing for safest corporations. Moves with risk-free rate + small credit spread

- **BAA10Y**

Moody’s BAA – 10Y Treasury Spread.Difference between BAA corporate yield and 10Y Treasury. Measures credit risk premium. Widens during stress/recessions

### Inflation & Prices

- **PCEPILFE**

Core PCE Price Index (Fed’s preferred inflation measure).Excludes food & energy. Used directly in Fed policy decisions

- **CPIAUCSL**

Consumer Price Index (All Urban Consumers). Headline inflation measure. Most cited inflation metric in media

### Labor Market

- **UNRATE**

Unemployment Rate, % of labor force unemployed

- **PAYEMS**

An initial claim is a claim filed by an unemployed individual after a separation from an employer. The claim requests a determination of basic eligibility for the Unemployment Insurance program.

- **AHETPI**

Average Hourly Earnings (Total Private). Wage growth indicator. Important for inflation pressure analysis

### Financial Conditions

- **NFCI**

Chicago Fed National Financial Conditions Index. Composite measure of financial stress. Includes credit spreads, volatility, leverage. Negative = loose conditions. Positive = tight/stressed conditions

- **USREC**

Binary recession indicator.

- **real_regime**

Categorical fed policy indicator based on real_rate = lending rate - inflation rate. Categories are divided into Easing, Tightening and Neutral periods defined by the the standard deviation above or below the mean.

## Data Processing:

The data comes in different granularities, it could be monthly, daily, or weekly depending on the explanatory variable. In order to work with a data set of at least $10^3$ we convert the data to weekly and split it into train test validate series before moving forward to the modelling part.
The actions taken are shown in the code below.

### Sourcing the data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE,
	cache = FALSE
)
options(scipen = 999)
```

```{r,include=TRUE}
#Store packages needed, in a vector
packages <- c("fredr", "dplyr", 
              "ggplot2","tidyr",
              "zoo","lubridate",
              "purrr","e1071",
              "patchwork","broom",
              "leaps","tidymodels",
              "rsample","lintr")
#Loop over every package, check if it exists otherwise install and load
for (package in packages)
  if(!requireNamespace(package, quietly = TRUE))
  {install.packages(package)}
  {library(package, character.only = TRUE)}
    require(package)
invisible(lapply(packages, library, character.only = TRUE))
```


```{r, include=FALSE,eval=FALSE}
#API key for FRED !gitignore
apikey<-"bfca9e4247ebe4a4f0915bb70b765d81"
```

```{r,eval=FALSE}
fredr_set_key(apikey) #set the api key for calling data from fred
```
- Defining a list of granularities made out of three lists of variables

***If adding more variables check the interval and add them to the correct list before rerunning***
```{r,eval=FALSE} 
symbols_monthly <- c("PCEPILFE", "CPIAUCSL", "UNRATE", "AHETPI","FEDFUNDS","USREC","PAYEMS")
symbols_weekly<- c("NFCI")
symbols_daily<- c("DGS2", "DGS10", "BAMLC0A1CAAAEY", "BAA10Y")
symbols<-list(monthly=symbols_monthly,daily=symbols_daily,weekly=symbols_weekly)
start_date<-as.Date("2000-01-01")
end_date<-as.Date("2026-02-20")#Variable assignments for our loop
```

- Creating nested series of list with value|data combination of different timeframes.
```{r,eval=FALSE}
df_l<-list(monthly = list(), weekly = list(), daily = list()) #create an empty nested list.
for (freq in names(symbols)) #loop over daily monthly weekly. 
  {for (sym in symbols[[freq]]) # inner loop over each independent variable.
    {df_l[[freq]][[sym]] <- fredr(
      series_id = sym,
      observation_start = start_date,
      observation_end = end_date
    ) %>% select(date, value) %>% rename(!!sym := value)}} #filter date, value and rename value column with the string contained in sym for each loop.
```

- We have our variables, now they need to be converted to weekly time frame format.

```{r,eval=FALSE}
weekly_monthly <- purrr::map(df_l$monthly, function(monthly_df) {

  # Create Friday-aligned weekly sequence
  weekly_dates <- tibble(
    date = seq(min(monthly_df$date),
               max(monthly_df$date),
               by = "week")
  ) |>
    mutate(date = floor_date(date, "week", week_start = 1) + 4)

  monthly_df |>
    right_join(weekly_dates, by = "date") |>
    arrange(date) |>
    tidyr::fill(everything(), .direction = "down")
})
head(weekly_monthly)
```

```{r,eval=FALSE}
weekly_daily <- lapply(df_l$daily, function(df) {
  df |>
    mutate(date = lubridate::floor_date(date, "week", week_start = 1) + 4)|>
    group_by(date) |>
    summarise(across(where(is.numeric), mean, na.rm = TRUE),
              .groups = "drop")
})
head(weekly_daily)
```

- Joining all series to a single dataset. 
```{r,eval=FALSE}
all_weekly_lists <- c(weekly_daily,
                      weekly_monthly,
                      df_l$weekly)

weekly_data <- reduce(all_weekly_lists,
                      full_join,
                      by = "date") |>
  arrange(date)
```
- Checking missing values due to shifting from monthly and daily to weekly values.
```{r,eval=FALSE}
sapply(weekly_data, function(x) sum(is.na(x)))
```
- Aligning and combining the data set on the same time frame.
```{r,eval=FALSE}
weekly_data <- weekly_data |>
  filter(date >= max(sapply(c(weekly_daily,
                              weekly_monthly,
                              df_l$weekly),
                            \(df) min(df$date)))) |>
  filter(date <= min(sapply(c(weekly_daily,
                              weekly_monthly,
                              df_l$weekly),
                            \(df) max(df$date)))) |>
  tidyr::drop_na()
```
- All on a consistent Friday weekly index. 

```{r}

```
```{r,eval=FALSE} 
# Our binary value is a Recession or non Recession flag metric either 0 or 1.
weekly_data %>%
  summarise(
    recession_weeks = sum(USREC == 1, na.rm = TRUE),
    percent_recession = mean(USREC == 1, na.rm = TRUE) * 100
  )
```
- Now Lets create categorical variables which will be Fed Rate Regimes.
```{r,eval=FALSE}
df_rates <- weekly_data %>%
  mutate(
    real_rate = FEDFUNDS - PCEPILFE
  )
```

```{r,eval=FALSE}
df_rates <- df_rates %>%
  mutate(
    real_regime = case_when(
      real_rate >= mean(real_rate, na.rm = TRUE) + sd(real_rate, na.rm = TRUE) ~ "Tighten",
      real_rate <= mean(real_rate, na.rm = TRUE) - sd(real_rate, na.rm = TRUE) ~ "Easing",
      TRUE ~ 
"Neutral"
    ),
    # Make it an ordered factor
    real_regime = factor(real_regime, levels = c("Easing", "Neutral", "Tighten"), ordered = TRUE)
  )
```


- These were data pre-processing and cleaning steps. Which consisted of acquisition of data from the Federal Reserve Database in any given time-frame and transmuting it into the weekly timescale.

- Daily to weekly timescales are transformed based on the mean weekly value.

- Monthly to weekly is transformed with the “last observation carried forward” (LOCF) interpolation.

```{r,eval=FALSE}
saveRDS(df_rates,"df_rates.rds")
getwd()
```

# Data Transformation and Modeling. 

```{r}
#Loading the data.
df_rates <- readRDS("df_rates.rds")
```
- Target variable plot.
```{r}
df_rates <- df_rates
plot(df_rates$date, df_rates$FEDFUNDS,
     type = "l",
     xaxt = "n",   # suppress default x-axis
     main = "Dependent Variable",
     col  = "steelblue",
     xlab = "Date",
     ylab = "Federal Funds Rate")
      
axis.Date(1,
          at = seq(min(df_rates$date),
                   max(df_rates$date),
                   by = "6 months"),
          format = "%Y-%m")
```
- Explanatory variables plot in different scales.
```{r}
df_rates |>
  select(date, DGS10, DGS2, NFCI, UNRATE, BAMLC0A1CAAAEY,
         BAA10Y, PCEPILFE, CPIAUCSL, AHETPI, PAYEMS ) |>
  pivot_longer(-date) %>%
  
  ggplot(aes(date, value)) +
  geom_line(color = "steelblue") +
  
  facet_wrap(~name, scales = "free", ncol = 2) +
  labs(
  title = 'Independent Variables') +

  theme_minimal()
  
```

## Data Scaling and Transformations.

A log transform will be applied for skewed time series before scaling them effectively for modelling.
```{r}
# compute skewness for numeric columns only
skew_vals <- sapply(df_rates[sapply(df_rates, is.numeric)],
                    function(x) skewness(x, na.rm = TRUE))

# choose threshold (common choices: 1 or 2) #Choose 0 for the whole list of features.
threshold <- 0

# return column names that exceed threshold
skewed_cols <- names(skew_vals[abs(skew_vals) > threshold])

skewed_cols #This is also a way to see which which variables are in our feature list. 
```



```{r}
vars <- c("DGS2","UNRATE","CPIAUCSL","PAYEMS",
          "AHETPI","PCEPILFE","NFCI",
          "BAA10Y","BAMLC0A1CAAAEY","DGS10")
formula <- as.formula(
  paste("FEDFUNDS ~", paste(vars, collapse = " + "))
)
linear_model <- lm(formula, data = df_rates)

```
Check https://uc-r.github.io/model_selection for best lm variable subset selection.
```{r}
summary(linear_model)
```
- We are getting a quite High R squared value of 0.80. Meaning that 80% of the data is explained by the model. It is unclear whether this is a meaningful explanation.

This leads to the question of what is the 20% that is left unexplained, how do other models fit the data. 



```{r}
car::vif(linear_model) # Multi-colinearity which returns very bad outcomes next step is transform and refit.
```


```{r}
formula_nvalues<-c("date", "USREC", "PCEPILFE", "real_rate", "real_regime")
df_diff <- df_rates %>%
  arrange(date) %>%
  mutate(
    across(c(DGS2, DGS10, BAMLC0A1CAAAEY, BAA10Y, 
             CPIAUCSL, PCEPILFE, AHETPI, PAYEMS), 
           ~ log(.) - lag(log(.))),
    across(c(NFCI, UNRATE, FEDFUNDS), 
           ~ . - lag(.))
  ) %>%
  slice(-1) 
```
Some scaling adjustments
```{r}
df_diff <- df_diff %>%
  mutate(
    AHETPI  = AHETPI * 10,
    PAYEMS  = PAYEMS * 150,
    UNRATE  = UNRATE * 10
  )
```

```{r}
df_diff |>
  select(date, DGS10, DGS2, NFCI, UNRATE, BAMLC0A1CAAAEY,
         BAA10Y, PCEPILFE, CPIAUCSL, AHETPI, PAYEMS ) |>
  pivot_longer(-date) %>%
  
  ggplot(aes(date, value)) +
  geom_line(color = "steelblue") +
  
  facet_wrap(~name, scales = "free", ncol = 2) +
  labs(
  title = 'Independent Transformed Variables') +

  theme_minimal()
  
```


```{r}
df_numeric <- df_diff %>%
  select(where(is.numeric))#Filters for only numeric values.

best_subset <- regsubsets(FEDFUNDS ~ .-real_rate, 
                          data = df_numeric, 
                         nvmax=12,       # maximum number of variables to consider
                          method = "exhaustive")  # tries all combinations
```
Inspect results
```{r}
summary_best <- summary(best_subset)
```
Plot metrics to choose the best subset
```{r}
par(mfrow = c(1,2))
plot(summary_best$cp, xlab = "Number of Variables", ylab = "Cp", type = "b")
points(which.min(summary_best$cp), min(summary_best$cp), col = "red", cex = 1, pch = 20)

plot(summary_best$adjr2, xlab = "Number of Variables", ylab = "Adjusted R²", type = "b")
points(which.max(summary_best$adjr2), max(summary_best$adjr2), col = "red", cex = 1, pch = 20)

```
Extract variable names for the best subset
```{r}
#Best variables to maximize Adjusted R2
best_vars <- names(coef(best_subset, which.max(summary_best$adjr2)))
best_vars<-best_vars[-1]
best_vars
```

```{r}
#Creating a vector of "variable names +" to feed into the linear model
formula2 <- as.formula(
  paste("FEDFUNDS ~", paste(best_vars, collapse = " + "))
) 
#Fitting a new linear model with the best selection to minimise R2
lm_best_fit <- lm(formula2, data=df_diff)
```

```{r}
#format(summary(lm_best_fit)$coefficients[, 1:4], scientific = TRUE, digits = 4)
```

```{r}
#Checking the residual structure
par(mfrow = c(2, 2))
plot(lm_best_fit)
```
```{r}
# Quick fix - add regime interaction
df_diff <- df_diff %>%
  mutate(
    hiking_regime = ifelse(date >= as.Date("2022-03-01") & 
                           date <= as.Date("2023-12-01"), 1, 0)
  )

# Fit with interaction
formula4 <- as.formula(
  paste("FEDFUNDS ~", 
        paste(best_vars, collapse = " + "),
        "+ hiking_regime")
)

lm_regime <- lm(formula4, data = df_diff)
summary(lm_regime)

par(mfrow = c(2, 2))
plot(lm_regime)
```
```{r}
# Function for rolling window cross-validation
rolling_cv <- function(data, formula, window_size, horizon = 1) {
  
  n <- nrow(data)
  predictions <- vector("list", n - window_size - horizon + 1)
  actual_values <- numeric(n - window_size - horizon + 1)
  train_dates <- vector("list", n - window_size - horizon + 1)
  test_dates <- vector("list", n - window_size - horizon + 1)
  
  for (i in 1:(n - window_size - horizon + 1)) {
    # Define training window
    train_idx <- i:(i + window_size - 1)
    # Define test point (or horizon)
    test_idx <- i + window_size + horizon - 1
    
    # Train model
    train_data <- data[train_idx, ]
    test_data <- data[test_idx, , drop = FALSE]
    
    #model <- lm(formula2, data = train_data)
    
    # Predict
    pred <- predict(lm_best_fit, newdata = test_data)
    
    # Store results
    predictions[[i]] <- pred
    actual_values[i] <- test_data$FEDFUNDS
    train_dates[[i]] <- data$date[train_idx]
    test_dates[[i]] <- data$date[test_idx]
  }
  
  # Calculate metrics
  predictions_vec <- unlist(predictions)
  
  results <- data.frame(
    test_date = unlist(test_dates),
    actual = actual_values,
    predicted = predictions_vec,
    error = actual_values - predictions_vec,
    squared_error = (actual_values - predictions_vec)^2,
    abs_error = abs(actual_values - predictions_vec)
  )
  
  return(results)
}

```

```{r}
# Test multiple forecast horizons
horizons <- c(1, 4, 13, 26)  # 1 week, 1 month, 1 quarter, 6 months

multi_horizon_cv <- lapply(horizons, function(h) {
  results <- rolling_cv(df_diff, formula2, window_size = 260, horizon = h)
  results$horizon <- h
  return(results)
}) %>% bind_rows()
```

```{r}
# Compare performance by horizon
multi_horizon_cv %>%
  group_by(horizon) %>%
  summarise(
    RMSE = sqrt(mean(squared_error)),
    MAE = mean(abs_error),
    n_predictions = n()
  )
```

```{r}
# Plot errors by horizon
ggplot(multi_horizon_cv, aes(x = as.factor(horizon), y = abs_error)) +
  geom_boxplot() +
  labs(title = "Prediction Error by Forecast Horizon",
       x = "Weeks Ahead", y = "Absolute Error") +
  theme_minimal()
```
```{r}
library(tidymodels)
library(rsample)
library(lubridate)

# ===== Split data by date =====
df_diff_clean <- df_diff %>% drop_na()

# Split at 2025-01-01
train_data <- df_diff_clean %>%
  filter(date < as.Date("2025-01-01"))

test_data <- df_diff_clean %>%
  filter(date >= as.Date("2025-01-01"))

cat("Training data:", nrow(train_data), "observations\n")
cat("Testing data:", nrow(test_data), "observations\n")
cat("Training period:", min(train_data$date), "to", max(train_data$date), "\n")
cat("Testing period:", min(test_data$date), "to", max(test_data$date), "\n")

# ===== Fit model on all pre-2025 data =====
model_pre2025 <- lm(formula2, data = train_data)

# ===== Predict on 2025+ =====
predictions_2025 <- test_data %>%
  mutate(
    predicted = predict(model_pre2025, newdata = test_data),
    error = FEDFUNDS - predicted,
    abs_error = abs(error),
    squared_error = error^2
  )

# ===== Calculate metrics =====
metrics_2025 <- predictions_2025 %>%
  summarise(
    RMSE = sqrt(mean(squared_error)),
    MAE = mean(abs_error),
    Bias = mean(error),
    Max_Error = max(abs_error),
    R2 = cor(FEDFUNDS, predicted)^2,
    n_predictions = n()
  )

print(metrics_2025)

# ===== Visualize actual vs predicted for 2025+ =====
ggplot(predictions_2025, aes(x = date)) +
  geom_line(aes(y = FEDFUNDS, color = "Actual"), linewidth = 1) +
  geom_line(aes(y = predicted, color = "Predicted"), linewidth = 1) +
  geom_ribbon(aes(ymin = predicted - 2*metrics_2025$RMSE, 
                  ymax = predicted + 2*metrics_2025$RMSE),
              alpha = 0.2, fill = "blue") +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "blue")) +
  labs(title = "2025+ Predictions vs Actual",
       subtitle = paste("RMSE:", round(metrics_2025$RMSE, 4), 
                       "| MAE:", round(metrics_2025$MAE, 4)),
       x = "Date",
       y = "FEDFUNDS (differenced)",
       color = NULL) +
  theme_minimal() +
  theme(legend.position = "top")

# ===== Plot prediction errors =====
ggplot(predictions_2025, aes(x = date, y = error)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") +
  geom_hline(yintercept = c(-2*metrics_2025$RMSE, 2*metrics_2025$RMSE),
             color = "blue", linetype = "dotted", alpha = 0.5) +
  labs(title = "Prediction Errors for 2025+",
       x = "Date",
       y = "Error (Actual - Predicted)") +
  theme_minimal()

# ===== Plot absolute errors =====
ggplot(predictions_2025, aes(x = date, y = abs_error)) +
  geom_line() +
  geom_point() +
  geom_hline(yintercept = metrics_2025$MAE, 
             color = "red", linetype = "dashed") +
  labs(title = "Absolute Prediction Errors for 2025+",
       subtitle = paste("Mean Absolute Error:", round(metrics_2025$MAE, 4)),
       x = "Date",
       y = "Absolute Error") +
  theme_minimal()

# ===== Identify worst predictions =====
worst_predictions <- predictions_2025 %>%
  arrange(desc(abs_error)) %>%
  select(date, FEDFUNDS, predicted, error, abs_error) %>%
  head(10)

cat("\n=== Worst 10 Predictions ===\n")
print(worst_predictions)

# ===== Summary table =====
predictions_2025 %>%
  select(date, FEDFUNDS, predicted, error) %>%
  print(n = Inf)  # Print all rows
```
```{r}
# 1. Check the raw FRED data for that specific date
df_rates %>%
  filter(date >= as.Date("2025-07-01") & date <= as.Date("2025-08-31")) %>%
  select(date, FEDFUNDS, PCEPILFE, CPIAUCSL, PAYEMS) %>%
  View()

# 2. If it's a data error, remove it and re-evaluate
predictions_2025_clean <- predictions_2025 %>%
  filter(date != as.Date("2025-08-01"))

# 3. Check if this affects your conclusions
cat("Model performs nearly perfectly on all 2025 data except one anomalous observation\n")
cat("This suggests either:\n")
cat("  a) Excellent model fit for normal conditions\n")
cat("  b) A data quality issue on Aug 1, 2025\n")
```
```{r}
# ===== Forecast each predictor variable =====
library(forecast)

# Function to forecast a variable
forecast_variable <- function(data, var_name, h = 8) {
  ts_data <- ts(data[[var_name]], frequency = 52)  # Weekly data
  
  # Use auto.arima for automatic forecasting
  model <- auto.arima(ts_data, seasonal = FALSE)
  forecasts <- forecast(model, h = h)
  
  return(as.numeric(forecasts$mean))
}

# Train on all data
train_full <- df_diff %>% 
  drop_na() %>%
  filter(date <= as.Date("2025-12-31"))

# Create future dates
future_dates <- tibble(
  date = seq(max(train_full$date) + 7, by = "week", length.out = 8)
) %>%
  mutate(date = floor_date(date, "week", week_start = 1) + 4)

# Forecast each predictor
future_predictors <- future_dates

for (var in best_vars) {
  future_predictors[[var]] <- forecast_variable(train_full, var, h = 8)
}

print(future_predictors)

# ===== Make final predictions =====
model_final <- lm(formula2, data = train_full)

predictions_2026_forecast <- future_predictors %>%
  mutate(
    predicted_FEDFUNDS = predict(model_final, newdata = future_predictors)
  )

predictions_2026_forecast %>%
  select(date, predicted_FEDFUNDS) %>%
  print(n = Inf)


```
```{r}
# 1. Get the last actual level before the forecast period
last_actual_date <- max(train_full$date)   # e.g., 2025-11-28 (Friday)
last_actual <- df_rates %>%
  filter(date == last_actual_date) %>%
  pull(FEDFUNDS)

# 2. Reverse difference the predictions
predictions_2026_level <- predictions_2026_forecast %>%
  arrange(date) %>%                               # ensure correct order
  mutate(
    # Cumulative sum of predicted differences
    cum_diff = cumsum(predicted_FEDFUNDS),
    # Add to last actual level
    FEDFUNDS_level = last_actual + cum_diff
  )

# 3. View the level forecasts
predictions_2026_level %>%
  select(date, predicted_FEDFUNDS, FEDFUNDS_level) %>%
  print(n = Inf)
```


# Appendix

## Bibliography/Sources

[FRED](https://fred.stlouisfed.org/)

[Linear Model Selection](https://uc-r.github.io/model_selection)

## Use of Generative AI

- It has been used for code syntax related questions since different coding approaches and packages are used throughout the document
- It has been also used in interpreting outputs and data transformation suggestions as well as modelling related approaches.
- I had to be critical about evaluating the function of suggested by AI lines of code. So i understand what they do and how efficiently they execute both in terms of performance and of syntax conciseness. 
